---
title: "stac_r_tutorial"
author: "Sam"
output: html_document
description: "A tutorial on using the STAC API with R, and opening Parquet datasets from S3."
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
# Install required packages if not already installed
if (!requireNamespace("rstac", quietly = TRUE)) {
  install.packages("rstac")
}
if (!requireNamespace("sf", quietly = TRUE)) {
  install.packages("sf")
}
if (!requireNamespace("terra", quietly = TRUE)) {
  install.packages("terra")
}
if (!requireNamespace("arrow", quietly = TRUE)) {
  install.packages("arrow")
}
if (!requireNamespace("leaflet", quietly = TRUE)) {
  install.packages("leaflet")
}
if (!requireNamespace("DT", quietly = TRUE)) {
  install.packages("DT")
}
if (!requireNamespace("dplyr", quietly = TRUE)) {
  install.packages("dplyr")
}
if (!requireNamespace("leaflet.extras", quietly = TRUE)) {
  install.packages("leaflet.extras")
}

library(rstac)
library(sf)
library(terra)
library(arrow)
library(leaflet)
library(DT)
library(dplyr)
library(leaflet.extras)
```

## Querying the STAC API

This script demonstrates how to query a STAC API and download data.

```{r stac-query-collections}
# Define the root STAC API endpoint
stac_endpoint <- "https://api.dive.edito.eu/data/"

# Query the root STAC API to get collections
collections_query <- stac(stac_endpoint) %>%
  rstac::collections() %>%
  get_request()

# Display the number of collections
cat("Number of collections:", length(collections_query$collections), "\n")
```

```{r stac-query-occurrence}
# Filter collections with 'occurrence' in their name or description directly
occurrence_collections <- Filter(function(collection) {
  grepl("occurrence", collection$title, ignore.case = TRUE)
}, collections_query$collections)

# Display the occurrence collections
print(occurrence_collections)

# Check if any matching collections were found
if (length(occurrence_collections) > 0) {
  occurrence_collection <- occurrence_collections[[1]]$id
  cat("Selected occurrence collection ID:", occurrence_collection, "\n")
} else {
  cat("No collections with 'occurrence' found.\n")
}
```

```{r stac-query-items}
# Query the items in the occurrence collection
if (exists("occurrence_collection")) {
  items_query <- stac(stac_endpoint) %>%
    stac_search(collections = occurrence_collection) %>%
    get_request()

  # Display the items query results
  print(items_query)

  # Extract and display assets from the first item
  if (length(items_query$features) > 0) {
    first_item <- items_query$features[[1]]
    print(first_item$assets)
  } else {
    cat("No items found in the query.\n")
  }
} else {
  cat("Occurrence collection not found.\n")
}
```


# Define a function to open a Parquet dataset from an S3 URL
```{r open-parquet function}

open_my_parquet <- function(s3_url) {
  # Parse the S3 URL to extract the endpoint and path
  temp <- strsplit(s3_url, "//")[[1]][2]  # Remove 'https://'
  endpoint <- strsplit(temp, "/")[[1]][1]  # Extract endpoint
  data_path <- sub(paste0(endpoint, "/"), "", temp)  # Extract path

  # Create the S3FileSystem connection
  data_lake <- S3FileSystem$create(anonymous = TRUE, 
                                   scheme = "https",
                                   endpoint_override = endpoint)

  # Try to open the dataset and handle any errors
  dataset <- NULL
  tryCatch({
    dataset <- arrow::open_dataset(data_path, filesystem = data_lake, format = "parquet")
  }, error = function(e) {
    cat("Error loading dataset:", e$message, "\n")
  })

  # Return the dataset (or NULL if loading failed)
  return(dataset)
}
```


# Extract the first item from the items query
```{r get-parquet}
if (length(items_query$features) > 0) {
  first_item <- items_query$features[[1]]

  # Show all assets in the first item
  print(first_item$assets)

  # Get the Parquet asset
  parquet_asset <- first_item$assets[["parquet"]]$href

  if (!is.null(parquet_asset)) {
    # Open the Parquet asset directly from S3 using the arrow package
    options(timeout = 120)
    parquet_data <- open_my_parquet(parquet_asset)

    if (!is.null(parquet_data)) {
      print(parquet_data$schema)
      print(head(parquet_data))
    } else {
      cat("Failed to load the Parquet dataset.\n")
    }
  } else {
    cat("No Parquet asset found in the first item.\n")
  }
} else {
  cat("No items found in the query.\n")
}
```

# Convert the Parquet dataset to a dataframe and display it
```{r open-parquet-dataframe-head}
# Convert the Parquet dataset to a dataframe and display the first 100,000 rows
if (!is.null(parquet_data)) {
  parquet_df <- as.data.frame(parquet_data[1:100000, ])

  # Display the first few rows of the dataframe
  print(head(parquet_df))
} else {
  cat("No Parquet dataset available to convert to a dataframe.\n")
}
```

# Query the 5 most frequent aphiaids and filter the dataset without loading into memory
```{r filter-aphiaids-efficient}
if (!is.null(parquet_data)) {
  # Use arrow's dplyr interface to count aphiaid frequencies directly
  aphiaid_counts <- parquet_data %>%
    group_by(aphiaid) %>%
    summarise(count = n()) %>%
    arrange(desc(count)) %>%
    collect()

  # Get the top 5 aphiaids
  top_aphiaids <- aphiaid_counts$aphiaid[1:5]

  # Filter the dataset for the 3rd, 4th, and 5th most frequent aphiaids
  filtered_data <- parquet_data %>%
    filter(aphiaid %in% top_aphiaids[3:5]) %>%
    collect()

  # Display the filtered dataframe
  print(head(filtered_data))
} else {
  cat("No Parquet dataset available to filter.")
}
```

# Update interactive table and map sections to use filtered_data
```{r display-interactive-table}
if (exists("filtered_data") && nrow(filtered_data) > 0) {
  # Sample a subset of the data if it is too large
  if (nrow(filtered_data) > 10000) {
    filtered_data <- filtered_data[sample(1:nrow(filtered_data), 10000), ]
  }

  # Display the filtered dataframe as an interactive table using the DT package
  datatable(
    filtered_data,
    options = list(
      pageLength = 10,
      scrollX = TRUE
    )
  )
} else {
  cat("No filtered dataset available to display as a table.\n")
}
```

```{r display-geometries-leaflet-filtered}
# if (exists("filtered_data") && nrow(filtered_data) > 0) {
#   # Check if the dataframe contains geometry information
#   if ("geometry" %in% colnames(filtered_data)) {
#     # Detect and decode geometry column
#     if (is.character(filtered_data$geometry)) {
#       # If geometry is in WKT format, use it directly
#       sf_data <- st_as_sf(filtered_data, wkt = "geometry")
#     } else if (inherits(filtered_data$geometry[[1]], "raw")) {
#       # If geometry is in WKB format, decode it
#       filtered_data$geometry <- lapply(filtered_data$geometry, function(x) {
#         st_as_text(st_as_sfc(list(x), EWKB = TRUE))
#       })
#       sf_data <- st_as_sf(filtered_data, wkt = "geometry")
#     } else {
#       stop("Unsupported geometry format.")
#     }

#     # Simplify geometries to reduce memory usage
#     sf_data <- st_simplify(sf_data, dTolerance = 0.01)

#     # Sample a subset of the data if it is too large
#     if (nrow(sf_data) > 10000) {
#       sf_data <- sf_data[sample(1:nrow(sf_data), 10000), ]
#     }

#     # Create a Leaflet map
#     leaflet_map <- leaflet(sf_data) %>%
#       addTiles()

#     # Add polygons if they exist
#     if (any(st_geometry_type(sf_data) %in% c("POLYGON", "MULTIPOLYGON"))) {
#       leaflet_map <- leaflet_map %>%
#         addPolygons(data = sf_data, color = "blue", weight = 1, fillOpacity = 0.5)
#     }

#     # Add points if they exist
#     if (any(st_geometry_type(sf_data) %in% c("POINT", "MULTIPOINT"))) {
#       leaflet_map <- leaflet_map %>%
#         addCircleMarkers(data = sf_data, radius = 5, color = "red")
#     }

#     # Print the map
#     print(leaflet_map)
#   } else {
#     cat("No geometry column found in the filtered dataset.\n")
#   }
# } else {
#   cat("No filtered dataset available to display geometries.\n")
# }
```

# Display geometries on a Leaflet map with interactive features
```{r display-geometries-leaflet-interactive}
if (exists("filtered_data") && nrow(filtered_data) > 0) {
  # Check if the dataframe contains geometry information
  if ("geometry" %in% colnames(filtered_data)) {
    # Detect and decode geometry column
    if (is.character(filtered_data$geometry)) {
      # If geometry is in WKT format, use it directly
      sf_data <- st_as_sf(filtered_data, wkt = "geometry")
    } else if (inherits(filtered_data$geometry[[1]], "raw")) {
      # If geometry is in WKB format, decode it
      filtered_data$geometry <- lapply(filtered_data$geometry, function(x) {
        st_as_text(st_as_sfc(list(x), EWKB = TRUE))
      })
      sf_data <- st_as_sf(filtered_data, wkt = "geometry")
    } else {
      stop("Unsupported geometry format.")
    }

    # Simplify geometries to reduce memory usage
    sf_data <- st_simplify(sf_data, dTolerance = 0.01)

    # Sample a subset of the data if it is too large
    if (nrow(sf_data) > 10000) {
      sf_data <- sf_data[sample(1:nrow(sf_data), 10000), ]
    }

    # Separate polygon and point geometries
    polygon_data <- sf_data[st_geometry_type(sf_data) %in% c("POLYGON", "MULTIPOLYGON"), ]
    point_data <- sf_data[st_geometry_type(sf_data) %in% c("POINT", "MULTIPOINT"), ]

    # Create a Leaflet map with interactive features
    leaflet_map <- leaflet() %>%
      addTiles()

    # Add polygons if they exist
    if (nrow(polygon_data) > 0) {
      leaflet_map <- leaflet_map %>%
        addPolygons(data = polygon_data, color = "blue", weight = 1, fillOpacity = 0.5, group = "Polygons")
    }

    # Add points if they exist
    if (nrow(point_data) > 0) {
      leaflet_map <- leaflet_map %>%
        addCircleMarkers(data = point_data, radius = 5, color = "red", group = "Points")
    }

    # Add interactive features
    leaflet_map <- leaflet_map %>%
      addLayersControl(
        overlayGroups = c("Polygons", "Points"),
        options = layersControlOptions(collapsed = FALSE)
      ) %>%
      addDrawToolbar(
        targetGroup = "Selected",
        editOptions = editToolbarOptions(selectedPathOptions = selectedPathOptions())
      ) %>%
      addMeasure(
        position = "topright",
        primaryLengthUnit = "meters",
        primaryAreaUnit = "sqmeters",
        activeColor = "#3D535D",
        completedColor = "#7D4479"
      )

    # Print the map
    print(leaflet_map)
  } else {
    cat("No geometry column found in the filtered dataset.\n")
  }
} else {
  cat("No filtered dataset available to display geometries.\n")
}
```

# Diagnostic step to check for map rendering issues
```{r diagnostic-check, echo=TRUE}
if (exists("filtered_data")) {
  cat("Filtered data exists with", nrow(filtered_data), "rows.\n")
} else {
  cat("Filtered data does not exist.\n")
}

if (exists("sf_data")) {
  cat("sf_data exists with", nrow(sf_data), "rows.\n")
} else {
  cat("sf_data does not exist.\n")
}

if (!requireNamespace("leaflet", quietly = TRUE)) {
  cat("Leaflet package is not installed.\n")
}

if (!requireNamespace("leaflet.extras", quietly = TRUE)) {
  cat("Leaflet.extras package is not installed.\n")
}
```